Type theories as quotient inductive-recursive types

### 2025-02-11 tex/main.tex
S2.4 on rewrite rules

We have formally implemented our definitions and proofs in \Agda, using the options \texttt{-{}-with-K}, \texttt{-{}-exact-split}, \texttt{-{}-rewriting}, and \texttt{-{}-local-confluence-check}.
Quotient inductive types are simulated by postulating equations to introduce equality constructors~\cite{Licata2011}.
Some of our statements are meta-theoretical and thus not checked by \Agda.
Others are only partially checked, as \Agda does not generate proof obligations for postulated quotient inductive types directly.
We have marked such `informal' proofs with the symbol \resizebox*{!}{\baselineskip}{\textdbend}.

@@

A key requirement for definitions with overlapping patterns is that, in addition to termination and completeness, any clause must eventually compute to the same term, i.e.\ $f$ is confluent.
Cockx shows~\cite[Proposition~2]{Cockx2014} that if a definition of $f$ is
\begin{enumerate}
  \item \emph{structurally recursive}, i.e.\ arguments of recursive calls are structurally smaller, and
  \item \emph{locally confluent}, i.e.\ for any two function clauses $f\;\overrightarrow{p_1} \reduce t_1$ and $f\;\overrightarrow{p_2} \reduce t_2$ with a most general unifer $\sigma$ for $\overrightarrow{p_1}$ and $\overrightarrow{p_2}$, their right-hand sides $\sigma t_1$ and $\sigma t_2$ are strictly equal, 
\end{enumerate}
then $f$ is confluent---essentially a variant of Newman's Lemma~\cite{Newman1942}.

\subsection{Definitions by rewrite rules}
\cite{Cockx2020,Cockx2021}

Unfortunately, neither general schemata of quotient inductive-inductive-recursive types nor definitions by overlapping patterns have been developed or implemented in existing proof assistants.

\paragraph*{Local confluence}
\LT{Explain the notion of critical pair; local peak}
\paragraph*{Strong normalisation}
\cite{Abel2002}
\LT{Explain how these two conditions imply global confluence}

>>>>

We have formally implemented our definitions and proofs in \Agda, using the options \texttt{-{}-with-K}, \texttt{-{}-exact-split}, \texttt{-{}-rewriting}, and \texttt{-{}-local-confluence-check}.
Quotient inductive types are simulated by postulating equations to introduce equality constructors~\cite{Licata2011}.
Some of our statements are meta-theoretical and thus not checked by \Agda.
Others are only partially checked, as \Agda does not generate proof obligations for postulated quotient inductive types.
We have marked such `informal' proofs with the symbol \resizebox*{!}{\baselineskip}{\textdbend}.

@@

A key requirement for definitions with overlapping patterns is that, in addition to termination and completeness, any clause must eventually compute to the same term, i.e.\ $f$ is confluent.
Cockx shows~\cite[Proposition~2]{Cockx2014} that if a definition of $f$ is
\begin{enumerate}
  \item \emph{structurally recursive}, i.e.\ arguments of recursive calls are structurally smaller, and
  \item \emph{locally confluent}, i.e.\ for any two function clauses $f\;\overrightarrow{p_1} \reduce t_1$ and $f\;\overrightarrow{p_2} \reduce t_2$ with a most general unifer $\sigma$ for $\overrightarrow{p_1}$ and $\overrightarrow{p_2}$, terms $\sigma t_1$ and $\sigma t_2$ have the same normal form, 
\end{enumerate}
then $f$ is confluent---essentially a variant of Newman's Lemma~\cite{Newman1942}.
We call a diagram $\sigma t_2 \Leftarrow f\;\sigma\overrightarrow{p_1} \Rightarrow \sigma t_1$ a \emph{local peak} informally.

\subsection{Overlapping patterns via adding rewrite rules}
Unfortunately, definitions by overlapping patterns are only implemented in an outdated experimental fork of \Agda.
To emulate a function by overlapping patterns in \Agda, we declare each function clause an additional rewrite rules \cite{Cockx2020,Cockx2021} (via the option \texttt{-{}-rewriting}) and use the local confluence checker (via \texttt{-{}-local-confluence-check}).
This is indeed one of the main use cases of user-defined rewrite rules discussed by Cockx~\cite{Cockx2020}. 
The caveat, though, is that the termination checker is not imposed, so we have to check the termination informally.
